A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-14 paper.Rmd
quick sketch of abstract

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Abstract to fill.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

In practice, over-interpretation of a single plot is common.  For instance, in @roy_chowdhury_using_2015, described an published example where the authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis. @roy_chowdhury_using_2015 showed that there were no differences in the expression levels between the gene groups. The lineup protocol proposed in @buja_statistical_2009 provides a solution to over-interpretation.  @majumder_validation_2013 showed that its use is analogous to the null hypothesis significance testing (NHST) framework. More specifically, the protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, have the identical graphical procedure as the data plot except the data is replaced with a data generation mechanism that is consistent with the null hypothesis, $H_0$. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires apriori specification of $H_0$ (or at least a null data generating mechanism) much like the requirement of knowing the distribution of the test statistic in NHST.

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the existence of numerous hypothesis test procedures. This paper provides evidence for why this is good advice, using data from a visual inference experiment. We show how conventional tests are too sensitive to be practically useful, and how the lineup protocol can be used to yield reliable and consistent reading of residual plots for better practical application. 
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

In practice, over-interpretation of a single plot is common.  For instance, @roy_chowdhury_using_2015, described a published example where the authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis. @roy_chowdhury_using_2015 showed that there were no differences in the expression levels between the gene groups. The lineup protocol proposed in @buja_statistical_2009 provides a solution to over-interpretation.  @majumder_validation_2013 showed that its use is analogous to the null hypothesis significance testing (NHST) framework. More specifically, the protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, have the identical graphical procedure as the data plot except the data is replaced with a data generation mechanism that is consistent with the null hypothesis, $H_0$. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires apriori specification of $H_0$ (or at least a null data generating mechanism) much like the requirement of knowing the distribution of the test statistic in NHST.

### 2023-04-14 paper.tex
quick sketch of abstract

\begin{abstract}
Abstract to fill.
\end{abstract}

@@

In practice, over-interpretation of a single plot is common. For
instance, in \citet{roy_chowdhury_using_2015}, described an published
example where the authors over-interpreted separation between gene
groups from a two-dimensional projection of a linear discriminant
analysis. \citet{roy_chowdhury_using_2015} showed that there were no
differences in the expression levels between the gene groups. The lineup
protocol proposed in \citet{buja_statistical_2009} provides a solution
to over-interpretation. \citet{majumder_validation_2013} showed that its
use is analogous to the null hypothesis significance testing (NHST)
framework. More specifically, the protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) plots, referred to as the \emph{null plots}, have the
identical graphical procedure as the data plot except the data is
replaced with a data generation mechanism that is consistent with the
null hypothesis, \(H_0\). Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\). This protocol requires apriori
specification of \(H_0\) (or at least a null data generating mechanism)
much like the requirement of knowing the distribution of the test
statistic in NHST.

>>>>

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the existence of numerous hypothesis test procedures.
This paper provides evidence for why this is good advice, using data
from a visual inference experiment. We show how conventional tests are
too sensitive to be practically useful, and how the lineup protocol can
be used to yield reliable and consistent reading of residual plots for
better practical application.
\end{abstract}

@@

In practice, over-interpretation of a single plot is common. For
instance, \citet{roy_chowdhury_using_2015}, described a published
example where the authors over-interpreted separation between gene
groups from a two-dimensional projection of a linear discriminant
analysis. \citet{roy_chowdhury_using_2015} showed that there were no
differences in the expression levels between the gene groups. The lineup
protocol proposed in \citet{buja_statistical_2009} provides a solution
to over-interpretation. \citet{majumder_validation_2013} showed that its
use is analogous to the null hypothesis significance testing (NHST)
framework. More specifically, the protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) plots, referred to as the \emph{null plots}, have the
identical graphical procedure as the data plot except the data is
replaced with a data generation mechanism that is consistent with the
null hypothesis, \(H_0\). Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\). This protocol requires apriori
specification of \(H_0\) (or at least a null data generating mechanism)
much like the requirement of knowing the distribution of the test
statistic in NHST.
